Parsing Postfix log files

### 2007-07-01 doc/logparser.tex
Renamed rule_order to hits; added hits_total; updated paper appropriately. Now graphing rule hits.

    \item [rule\_order] is an efficiency measure.  This counter is
        maintained for every rule and incremented each time the rule
        successfully matches.  At the start of each run the program sorts
        the rules in descending rule\_order, and at the end of the run
        updates every rule's rule\_order.  Assuming that the distribution
        of log lines is reasonably consistent between log files, rules
        matching more commonly occurring log lines will be tried before
        rules matching less commonly occurring log lines, lowering the
        program's execution time.  Rule ordering for efficiency is
        discussed in section~\ref{rule ordering for efficiency}.

    \item [priority] This is the user-configurable companion to
        rule\_order: rules will be tried in order of priority, overriding
        rule\_order.  This allows more specific rules to take precedence
        over more general rules.

@@

The parser does not try to detect overlapping rules;\footnote{It may be
possible to parse each rule's regex, generate an NFA, and compare all NFAs
to determine if any overlap.  The author has not attempted to do this: such
a project would probably qualify for a PhD, and may involve solving the
Halting Problem and circumventing the Church-Turing Thesis.} that
responsibility is left to the author of the rules.  Unintentionally
overlapping rules lead to inconsistent parsing and data extraction because
the order in which rules are tried against each line determines which rule
matches.  Overlapping rules are frequently a requirement, allowing a more
specific rule to match some lines and a more general rule to match the
remainder, e.g.\ separating smtp delivery to a filtering proxy from smtp
delivery to the rest of the world.  The algorithm provides a facility for
handling overlapping rules: the priority field in each rule (defaults to
zero).  Rules are tried in order of priority, highest first, overriding the
rule\_order field maintained by the parser.  Negative priorities may also
be useful for catchall rules.

@@

\textbf{Field}      & \textbf{Value}                                    \\
name                & Unknown sender domain                             \\
description         & We do not accept mail from unknown domains        \\
restriction\_name   & reject\_unknown\_sender\_domain                   \\
postfix\_action     & REJECTED                                          \\
program             & postfix/smtpd                                     \\
regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
                    & \verb!Sender address rejected: Domain not found;! \\
                    & \verb!from=<\5> to=<(__RECIPIENT__)> !            \\
                    & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
result\_cols        & recipient = 6; sender = 5                         \\
connection\_cols    & helo = 7                                          \\
result\_data        &                                                   \\
connection\_data    &                                                   \\
action              & REJECTION                                         \\
queueid             & 1                                                 \\
rule\_order         & 0                                                 \\
priority            & 0                                                 \\

@@

@@

XXX DO A FULL RUN ONCE RULE\_ORDER ISN'T BEING TRASHED AND GRAPH RULE
HITS\@.

Rule ordering was briefly discussed in section~\ref{rule attributes} and
it will be covered in greater detail in this section.  At the time of
writing there are 131 different rules, some of which match much more
frequently than others.  Assuming that the distribution of log lines is
reasonably steady over time, program efficiency should benefit from trying
more frequently matching rules before those which match less frequently.
To test this hypothesis three full test runs were performed with different
rule orderings:

@@

    \item [MAIL\_PICKED\_FOR\_DELIVERY] This action represents Postfix
        picking a mail from the queue to deliver. This action is used for
        both qmgr and cleanup due to out of order log lines; see
        section~\ref{discarding-cleanup-lines} for details.

    \item [PICKUP] Pickup is the service which deals with mail submitted
        locally via /usr/sbin/sendmail. This action creates a new state
        table entry and saves data to it, unless out of order logging has
        caused the cleanup line to be logged first, as discussed in
        section~\ref{pickup-logging-after-cleanup}.

@@

This complication is further complicated by out of order cleanup lines: see
section~\ref{discarding-cleanup-lines} for details.

@@

\label{discarding-cleanup-lines}

@@

\label{pickup-logging-after-cleanup}

@@

Full coverage of log lines is both easy to achieve and hard to maintain.
It is easy to achieve full coverage for a limited set of log files (at the
time of writing the parser has 131 rules, fully parsing 93 contiguous log
files from Postfix 2.2 and 2.3), and new rules are easy to add.
Maintaining full coverage is hard because other servers have different
restrictions with custom messages, RBLs change their messages over time,
major releases of Postfix change warning messages (usually adding more
information), etc.\ so over time the log lines drift and change.

XXX GRAPH RULE HITS

@@

The first of these problems cannot be dealt with; although out of order log
lines (the second problem) are dealt with in other circumstances (see
sections~\ref{tracking-re-injected-mail}, \ref{discarding-cleanup-lines}
and \ref{pickup-logging-after-cleanup}), dealing with these lines would
introduce a large amount of complication (deferred warnings, deferred
committal, ambiguity) for a problem which occurs only four times in
60721709 log lines.

@@

>>>>

    \item [hits] is an efficiency measure.  This counter is maintained for
        every rule and incremented each time the rule successfully matches.
        At the start of each run the program sorts the rules in descending
        hits, and at the end of the run updates every rule's hits.
        Assuming that the distribution of log lines is reasonably
        consistent between log files, rules matching more commonly
        occurring log lines will be tried before rules matching less
        commonly occurring log lines, lowering the program's execution
        time.  Rule ordering for efficiency is discussed in
        section~\ref{rule ordering for efficiency}.

    \item [hits\_total] The total number of hits for this rule over all
        runs of the parser.

    \item [priority] This is the user-configurable companion to hits: rules
        will be tried in order of priority, overriding hits.  This allows
        more specific rules to take precedence over more general rules.

@@

The parser does not try to detect overlapping rules;\footnote{It may be
possible to parse each rule's regex, generate an NFA, and compare all NFAs
to determine if any overlap.  The author has not attempted to do this: such
a project would probably qualify for a PhD, and may involve solving the
Halting Problem and circumventing the Church-Turing Thesis.} that
responsibility is left to the author of the rules.  Unintentionally
overlapping rules lead to inconsistent parsing and data extraction because
the order in which rules are tried against each line determines which rule
matches.  Overlapping rules are frequently a requirement, allowing a more
specific rule to match some lines and a more general rule to match the
remainder, e.g.\ separating smtp delivery to a filtering proxy from smtp
delivery to the rest of the world.  The algorithm provides a facility for
handling overlapping rules: the priority field in each rule (defaults to
zero).  Rules are tried in order of priority, highest first, overriding the
hits field maintained by the parser.  Negative priorities may also be
useful for catchall rules.

@@

\textbf{Field}      & \textbf{Value}                                    \\
name                & Unknown sender domain                             \\
description         & We do not accept mail from unknown domains        \\
restriction\_name   & reject\_unknown\_sender\_domain                   \\
postfix\_action     & REJECTED                                          \\
program             & postfix/smtpd                                     \\
regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
                    & \verb!Sender address rejected: Domain not found;! \\
                    & \verb!from=<\5> to=<(__RECIPIENT__)> !            \\
                    & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
result\_cols        & recipient = 6; sender = 5                         \\
connection\_cols    & helo = 7                                          \\
result\_data        &                                                   \\
connection\_data    &                                                   \\
action              & REJECTION                                         \\
queueid             & 1                                                 \\
hits                & 0                                                 \\
hits\_total         & 0                                                 \\
priority            & 0                                                 \\

@@

    \item [hits, hits\_total and priority:] hits and priority are used in
        ordering the rules (see section~\ref{rule ordering for
        efficiency}); hits\_total is updated at the end of each run but is
        otherwise unused.

@@

Rule ordering was briefly discussed in section~\ref{rule attributes} and it
will be covered in greater detail in this section.  At the time of writing
there are 131 different rules, the majority of which match several orders
of magnitude less frequently than the top 10\%, as shown in graph~\ref{rule
hits}.  Assuming that the distribution of log lines is reasonably steady
over time, program efficiency should benefit from trying more frequently
matching rules before those which match less frequently.  To test this
hypothesis three full test runs were performed with different rule
orderings:

@@

    \item [MAIL\_PICKED\_FOR\_DELIVERY] This action represents Postfix
        picking a mail from the queue to deliver. This action is used for
        both qmgr and cleanup due to out of order log lines; see
        section~\ref{discarding cleanup lines} for details.

    \item [PICKUP] Pickup is the service which deals with mail submitted
        locally via /usr/sbin/sendmail. This action creates a new state
        table entry and saves data to it, unless out of order logging has
        caused the cleanup line to be logged first, as discussed in
        section~\ref{pickup logging after cleanup}.

@@

This complication is further complicated by out of order cleanup lines: see
section~\ref{discarding cleanup lines} for details.

@@

\label{discarding cleanup lines}

@@

\label{pickup logging after cleanup}

@@

Full coverage of log lines is both easy to achieve and hard to maintain.
It is easy to achieve full coverage for a limited set of log files (at the
time of writing the parser has 131 rules, fully parsing 93 contiguous log
files from Postfix 2.2 and 2.3), and new rules are easy to add.
Maintaining full coverage is hard because other servers have different
restrictions with custom messages, RBLs change their messages over time,
major releases of Postfix change warning messages (usually adding more
information), etc.\ so over time the log lines drift and change.
Graph~\ref{rule hits} shows the hits for each rule over all 93 log files;
it's obvious that a small number of rules match the vast majority of the
lines, and more than half the rules match fewer than 100 times.

@@

The first of these problems cannot be dealt with; although out of order log
lines (the second problem) are dealt with in other circumstances (see
sections~\ref{tracking-re-injected-mail},~\ref{discarding cleanup
lines}~and~\ref{pickup logging after cleanup}), dealing with these lines
would introduce a large amount of complication (deferred warnings, deferred
committal, ambiguity) for a problem which occurs only four times in
60721709 log lines.

@@

\subsubsection{Rule hits}
\label{rule hits}
\includegraphics{plot-hits.ps}
